---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# thermoclustr

<!-- badges: start -->
[![R-CMD-check](https://github.com/tobiste/thermoclustr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tobiste/thermoclustr/actions/workflows/R-CMD-check.yaml)
<!-- [![Codecov test coverage](https://codecov.io/gh/tobiste/thermoclustr/graph/badge.svg)](https://app.codecov.io/gh/tobiste/thermoclustr) -->
[![pkgdown](https://github.com/tobiste/thermoclustr/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/tobiste/thermoclustr/actions/workflows/pkgdown.yaml)
<!-- badges: end -->

The goal of {thermoclustr} is to provide tools to further analyze thermal history 
models in thermochronology, including estimating cooling path densities, and 
path families.

## Installation

You can install the development version of {thermoclustr} from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tobiste/thermoclustr")
```

## Example

The following minimal example shows how to import data, visualize the paths and the path density as well as how to filter, and cluster the t-T paths into 3 path families.

```{r import, message=FALSE, warning=FALSE}
library(thermoclustr)
library(dplyr)
library(ggplot2)

# load example dataset of a HeFTy model output
path2myfile <- system.file("112-73_30_H1_50-inv.txt", package = "thermoclustr")
tT_paths <- read_hefty(path2myfile)

# set `theme_classic()` as the default ggplot theme
theme_set(theme_classic())
```

The HeFTy model contains the modeled paths, the initial model constraints, 
the weighted mean path, and some summary statistics on the mineral grains. 
{thermochlustr} imports the model as a `list` that contains the aforementioned 
features as list entries.

Thus, to visualize the paths with need to extract the `path` object from the 
imported model `tT_paths`:

```{r plot}
# (Optional) Create a subset of interest of the data:
tT_paths_cropped <- crop_paths(tT_paths, time = c(0, 400), temperature = c(0, 250))

tT_paths_cropped$paths |>
  ggplot(aes(time, temperature, color = Comp_GOF, group = segment)) +
  geom_path() +
  scale_color_viridis_c("GOF") +
  scale_x_reverse(position = "top") +
  scale_y_reverse()
```

The path density can be visualized using `plot_path_density_filled()`:

```{r density}
plot_path_density_filled(tT_paths_cropped, geom = "raster") +
  scale_x_reverse(position = "top") +
  scale_y_reverse() +
  scale_fill_viridis_c()
```

To cluster the data, the following steps are required:

```{r cluster}
# Cluster the paths
paths_cluster <- cluster_paths(tT_paths_cropped, k = 3)

# Join with path dataset
paths_clustered <- merge(
  tT_paths_cropped$paths,
  paths_cluster,
  by = "segment"
)
```

Finally, the visualization of the clustered t-T paths:

```{r plot2}
paths_clustered |>
  ggplot(
    aes(
      x = time,
      y = temperature,
      color = cluster,
      group = segment
    )
  ) +
  geom_path() +
  scale_x_reverse(position = "top") +
  scale_y_reverse() +
  scale_color_viridis_d("Cluster", begin = .1, end = .9)
```
