---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# thermochron

<!-- badges: start -->
[![R-CMD-check](https://github.com/tobiste/thermochron/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tobiste/thermochron/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/tobiste/thermochron/branch/main/graph/badge.svg)](https://app.codecov.io/gh/tobiste/thermochron?branch=main)
<!-- badges: end -->

The goal of thermochron is to provide tools to further analyze thermal history 
models in thermochronology, including estimating cooling path densities, and 
path families.

## Installation

You can install the development version of thermochron from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tobiste/thermochron")
```

## Example

The following minimal example shows how to import data, visualize the paths and the path density as well as how to filter, and cluster the tT paths into 3 path families.


```{r import}
library(thermochron)
library(dplyr)
library(ggplot2)

# load example dataset of a HeFTy model output
path2myfile <- system.file("112-9-30-zr-inv.txt", package = "thermochron")
tT_paths <- read_hefty(path2myfile)
```

The HeFTy model contains the modeled paths, the initial model constraints, 
the weighted mean path, and some summary statistics on the mineral grains. 
{thermochron} imports the model as a `list` that contains the aforementioned 
features as list entries.

Thus, to visualize the paths with need to extract the `path` object from the 
imported model `tT_paths`:

```{r plot}
tT_paths$paths |> 
  ggplot(aes(time, temperature, color = Comp_GOF, group = segment)) +
  geom_path() +
  scale_color_viridis_c() +
  scale_x_reverse(position = "top") +
  scale_y_reverse()
```

The path density can be visualized using `plot_path_density_filled()`:

```{r density}
plot_path_density_filled(tT_paths, show.legend = FALSE) +
  scale_x_reverse(position = "top") +
  scale_y_reverse()
```

To cluster a subset of the data, the following steps are required:

```{r cluster}
# extract the paths and filter to time range of interest:
my_paths <- tT_paths$paths |> 
  dplyr::filter(time <= 500, temperature <= 200)

# cluster the paths
my_paths_cluster <- cluster_paths(my_paths, cluster = 3)

# Join with path dataset
my_paths_clustered <- dplyr::left_join(
  my_paths,
  my_paths_cluster,
  dplyr::join_by(segment)
)
```

Finally, the visualization of the clustered tT paths:

```{r plot2}
my_paths_clustered |>
  ggplot(
    aes(
      x = time, 
      y = temperature,
      color = cluster,
      alpha = Comp_GOF,
      group = segment
    )
  ) +
  geom_path() +
  scale_x_reverse(position = "top") +
  scale_y_reverse()
```
