---
title: "Cluster modelled t-T paths"
author: "Tobias Stephan"
date: "2025-01-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,dev = "ragg_png",
                      dpi = 400)
```

## R Markdown

```{r install, message=TRUE, warning=TRUE}
# remotes::install_github("tobiste/HeFTy.SmoothR", auth_token = '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') # comment this line out after installation
library(HeFTy.SmoothR)
library(tidyverse)
library(patchwork)
library(ggtext)
library(scico)
library(ggtext)
library(glue)
library(viridis)
library(systemfonts)
library(ggalt)

theme_set(theme_classic(base_family = "Roboto Condensed", base_size = 12))
theme_update(
  panel.border = element_rect(color = 'black', fill = NA, linewidth = rel(1)),
  axis.line = element_blank(),
  #text = element_markdown(),
  axis.title = element_markdown(color = 'grey10'),
  legend.title = element_text(color = 'grey10'), 
  plot.title.position = 'plot',
  plot.subtitle.position = 'plot',
  #legend.position = 'bottom',
  plot.title = element_markdown(face = 'bold', color = 'grey10'),
  plot.subtitle = element_markdown(color = 'darkgray', face = 'bold'),
  plot.caption  = element_markdown(color = 'darkgray'),
  strip.background = element_rect(fill = 'lightgrey', color = NA),
  strip.text.y = element_markdown(color = 'white', angle = 0, face = 'bold')
)
set.seed(20250411)
```

Import the data from file
```{r data_import, message=TRUE, warning=TRUE}
#fname <- "C:/Users/taisf/Desktop/112-9-30-zr-inv.txt" # change file path here
fname <- "C:/Users/tstephan/Downloads/112-9-30-zr-inv.txt" # change file path here
# fname <- file.choose() # uncomment this line to select the file interactively

my_data <- read_hefty(fname)
my_paths <- my_data$paths |> 
  arrange(Fit, Comp_GOF) |> 
  mutate(segment = fct_reorder(segment, Comp_GOF))

mean_path <- my_data$weighted_mean_path

constraints <- my_data$constraints
constraint_box <- do.call(rbind, lapply(seq_along(constraints), function(i) {
  data.frame(
    x = c(constraints$max_time[i], constraints$max_time[i], constraints$min_time[i], constraints$min_time[i]),
    y = c(constraints$max_temp[i], constraints$min_temp[i], constraints$min_temp[i], constraints$max_temp[i]),
    group = constraints$constraint[i]
  )
}))

```

```{r sample_name}
sample_name <- 'Sample' # change sample name here

use_viridis <- FALSE
viridis_option <- 'G' # options are 'A' for magma, 'B' for inferno, 'C' for plasma, 'D' for viridis, 'E' for cividis, 'F' for rocket, 'G' for mako, 'H' for turbo
sico_palette <- 'vanimo' 
```

Filter the data (time, and temperature range):

```{r filter_setting}
max_time <- 500
max_temperature <- 250
```

```{r filter}
my_data_filtered <- my_paths |>
  dplyr::filter(time <= max_time, temperature <= max_temperature) # change filter settings here
```

## Plot the paths 

```{r paths, message=TRUE, warning=TRUE}
add_constraints <- TRUE
add_mean_path <- TRUE

plot1 <- ggplot(my_data_filtered, aes(time, temperature, group = segment, color = Comp_GOF)) +
  geom_path() +
  labs(
    title = "Modelled t-T paths",
    x = "**Time** (Ma)",
    y = "**Temperature** (&deg;C)"
  ) +
  {
    if (add_mean_path) {
      geom_path(
        data = mean_path, aes(time, temperature),
        color = "black", linewidth = 1, linetype = "dashed", inherit.aes = FALSE
      )
    }
  } +
  {
    if (add_constraints) {
      geom_polygon(
        data = constraint_box, aes(x, y, group = group),
        fill = NA, color = "black", linetype = "dotted", lwd = .9, inherit.aes = FALSE
      )
    }
  } +
  {
    if (use_viridis) {
      scale_color_viridis_b(
        "Goodnes-of-fit (GOF)",
        limits = c(0, .51),
        breaks = seq(0, 1, .1),
        oob = scales::squish,
        option = viridis_option
      )
    } else {
      scale_color_scico(
        "Goodnes-of-fit (GOF)",
        palette = sico_palette,
        limits = c(0, .51),
        breaks = seq(0, 1, .1),
        oob = scales::squish,
      )
    }
  } +
  scale_x_reverse(position = "top") +
  scale_y_reverse() +
  guides(color = guide_colourbar(
    title.position = "top", 
    title.hjust = .5, 
    barwidth = unit(20, "lines"), 
    barheight = unit(.5, "lines"))
    ) +
  theme(
    legend.position = "bottom"
  ) +
  coord_cartesian(expand = FALSE, ylim = c(max_temperature, 0), xlim = c(max_time, 0))
print(plot1)
```

## Plot the path density

```{r density}
plot2 <- my_data_filtered |>
  plot_path_density_filled(GOF_rank = Inf) +
  labs(
    title = "Path kernel density",
    x = "**Time** (Ma)",
    y = "**Temperature** (&deg;C)"
  ) +  
  {if (use_viridis){
    scale_fill_viridis_d(option = viridis_option) 
  }
    else {
      scale_fill_scico_d(palette = sico_palette)
    }
    } +
  scale_x_continuous(transform = "reverse", position = "top") +
  scale_y_continuous(transform = "reverse") +
  coord_cartesian(expand = FALSE, ylim = c(max_temperature, 0), xlim = c(max_time, 0)) +
  guides(fill = "none")

print(plot2)
```



## Cluster the paths

Enter cluster settings here:
```{r cluster_set}
cluster_method <- "kmeans" # options are "kmeans", "hclust", and "gam"
number_of_clusters <- 3
```

Perform the clustering using the above settings and calculate binned statistics:
```{r cluster}
data_cluster <- my_data_filtered |>
  cluster_paths(cluster = number_of_clusters, method = cluster_method) |>
  group_by(cluster) |>
  mutate(n = n()) |>
  right_join(my_data_filtered, join_by(segment))

data_cluster_binned <- densify_cluster(data_cluster) |> 
  path_statistics()
```

### Plot the cluster modelled t-T paths

Plot settings:
```{r plot_settings}
show_bars <- TRUE
show_median_lines <- TRUE

use_default_colors <- TRUE # if true colors will be selected automatically from the viridis palette. if not, colors will be picked from color specification below.

# Specify your custom colors for each clusters here:
my_cluster_colors <- c(
  '1' = 'darkred',
  '2' = 'dodgerblue',
  '3' = 'seagreen'
)

path_opacity <- .1
```

Plot the cluster modeled t-T paths:
```{r cluster_plot}
plot3 <- ggplot(data = data_cluster, aes(time, temperature, color = cluster, fill = cluster)) +
  {
    if (show_bars) {
      geom_ribbon(data = data_cluster_binned, 
                  aes(x = time_median,  ymin=temp_5, ymax=temp_95, group = cluster),
                  stat="stepribbon",
                  fill = "grey75", color = "grey90", inherit.aes = FALSE, alpha = 1
      )
    }
  } +
  geom_line(aes(group = segment), alpha = path_opacity, show.legend = FALSE) +
  {
    if (show_median_lines) { #
      geom_step(
        data = data_cluster_binned, aes(x = time_median, y = temp_median),
        lwd = 1.5, color = "grey20", show.legend = FALSE
      )
    }
  } +
  geom_text(
    data = data_cluster |> select(cluster, n) |> distinct(),
    aes(label = glue("N: {n}")),
    x = 0, y = -(max_temperature-20), color = "black", hjust = 1
  ) +
  
   # Color settings:
  {
    if (use_default_colors) {{
      if(use_viridis){
      scale_color_viridis_d(aesthetics = c("fill", "color"), end = .9, option = viridis_option)
      } else {
        scale_color_scico_d(palette = sico_palette, aesthetics = c("fill", "color"))
      }}
    } else {
      scale_color_manual(values = my_cluster_colors)
    }
  } +
  
  # Axes settings:
  scale_y_reverse() +
  scale_x_reverse(position = "top") +
  
  # Labels:
  labs(
    title = sample_name,
    subtitle = "Path clusters", 
    caption = glue("Cluster method: {cluster_method}"), 
    x = "**Time** (Ma)", 
    y = "**Temperature** (&deg;C)"
    ) +
  
  # Theme settings:
  
  theme(
    #strip.background = element_rect(fill = NA, linewidth = 0),
    #strip.placement = "outside",
    #strip.text.y = element_text(angle=0),
    panel.grid.major = element_line(color = 'grey')
    ) +
  facet_wrap(vars(cluster), ncol = 1, strip.position = "right") +
  coord_cartesian(expand = FALSE, ylim = c(max_temperature, 0), xlim = c(max_time, 0))

print(plot3)
```

## Summary plot
Plot all plots in one figure:
```{r summary}
print(plot1 + plot2 + plot3)
```

## Save plots
```{r save}
file_extension <- 'png' # change file format here
scale <- 0.5 # change font and line widths scaling factor here

path1 <- gsub('.txt', paste0('_paths.', file_extension), fname)
path2 <- gsub('.txt', paste0('_path_density.', file_extension), fname)
path3 <- gsub('.txt', paste0('_clusters.', file_extension), fname)

# width and height are in inches
ggsave(path1, plot1, width = 11, height = 8, scale = scale, dpi  = 600) # save paths
ggsave(path2, plot2, width = 11, height = 8, scale = scale, dpi  = 600) # save density
ggsave(path3, plot3, width = 11, height = 11, dpi = 600, scale = scale) # save clusters
```


### Single clusters plots

```{r single_clusters}
for (i in 1:number_of_clusters) {
  plot_cluster <- ggplot(data = data_cluster |> filter(cluster == i), aes(time, temperature, color = cluster, fill = cluster)) +
    {
    if (show_bars) { #
       geom_ribbon(data = data_cluster_binned |> filter(cluster == i), 
                  aes(x = time_median, ymin=temp_5, ymax=temp_95, group = cluster),
                  stat="stepribbon",
                  fill = "grey75", color = "grey90", inherit.aes = FALSE, alpha = 1
      )
    }
  } +
  geom_line(aes(group = segment), alpha = path_opacity, show.legend = FALSE) +
  {
    if (show_median_lines) { #
      geom_step(
        data = data_cluster_binned |> filter(cluster == i) , aes(x = time_median, y = temp_median),
        lwd = 1.5, color = "grey20", show.legend = FALSE
      )
    }
  } +
  geom_text(
    data = data_cluster  |> filter(cluster == i) |> select(cluster, n) |> distinct(),
    aes(label = paste0("N: ", n)),
    x = 0, y = -(max_temperature-10), color = "black", hjust = 1.1
  ) +
  
   # Color settings:
  {
    if (use_default_colors) {
      scale_color_viridis_d(aesthetics = c("fill", "color"), end = .9, option = viridis_option)
    } else {
      scale_color_manual(values = my_cluster_colors)
    }
  } +
  
  # Axes settings:
  scale_y_reverse() +
  scale_x_reverse(position = "top") +
  
  # Labels:
  labs(
    title = sample_name,
    subtitle = glue("Path cluster {i}"),
    caption = glue("Cluster method: {cluster_method}"), 
    x = "**Time** (Ma)", 
    y = "**Temperature** (&deg;C)"
    ) +
  
  # Theme settings:
  theme(
    panel.grid.major = element_line(color = 'grey')
    ) +
  coord_cartesian(expand = FALSE, ylim = c(max_temperature, 0), xlim = c(max_time, 0))
  
  path3_i <- gsub('.txt', paste0('_cluster-', i, '.', file_extension), fname)
  ggsave(path3_i, plot_cluster, width = 11, height = 8, scale = scale)
}
```

