---
title: "Path families"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#output: pdf_document
#latex_engine: xelatex 
vignette: >
  %\VignetteIndexEntry{Path families}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(thermochron)
library(ggplot2)
library(dplyr)
library(forcats)
library(scales)

# set a seed for reproducibility as some algorithms generate random data
set.seed(20250411)
```

This document provides a brief working example to demonstrate how to
cluster paths from HeFTy inverse thermal history models.

Load some example data:

```{r data}
data(tT_paths1)

# Extract the paths and reorder the paths by the GOF values:
paths <- tT_paths1$paths |>
  arrange(Fit, Comp_GOF) |>
  mutate(segment = fct_reorder(segment, Comp_GOF))
```


Extract the weighted mean path from the loaded objects, and build a constrain box from the min/max values given for the constraints (later used for plotting):

```{r data2}
# Extract mean path:
mean_path <- tT_paths1$weighted_mean_path

# Extract the constraints:
constraints <- tT_paths1$constraints

# build a constraint box polygon:
constraint_box <- do.call(rbind, lapply(seq_along(constraints), function(i) {
  data.frame(
    x = c(constraints$max_time[i], constraints$max_time[i], constraints$min_time[i], constraints$min_time[i]),
    y = c(constraints$max_temp[i], constraints$min_temp[i], constraints$min_temp[i], constraints$max_temp[i]),
    group = constraints$constraint[i]
  )
}))
```

You can quickly inspect your Hefty output by calling individual list elements:
```{r path_summary}
print(tT_paths1$summary)
```

To analyze a *subset* of the data, you can filter the data using a time and temperature range. For example:

```{r filter_setting}
max_time <- 400
max_temperature <- 250

paths_filtered <- crop_paths(
  tT_paths1,
  time = c(-Inf, max_time),
  temperature = c(-Inf, max_temperature)
) 
```

You may also filter the data by a GOF threshold or range. Maybe first you may check how the GOF values are distributed:

```{r gof_hist}
hist(paths_filtered$paths$Comp_GOF, 
     xlim = c(0, 1), 
     main = "Histogram of paths' goodness-of-fit statistics")
```

In our example, we will **not** filter the GOF values, because we hope the clustering may identify "hidden" path families that might go undetected if we filter the data too much. If you still want to filter the paths, you could do it likes this:

```{r filter_god, eval=FALSE, include=TRUE}
paths_filtered$paths <- subset(paths_filtered$paths, Comp_GOF > 0.2)
```

Now we *visualize* the filtered t-T paths:

```{r paths, message=TRUE, warning=TRUE, fig.dim = c(7, 5)}
ggplot(paths_filtered$paths, aes(time, temperature, group = segment, color = Comp_GOF)) +
  geom_path() +
  labs(
    title = "Modelled t-T paths",
    x = "Time (Ma)",
    y = bquote("Temperature (" * degree * "C)")
  ) +
  geom_path(
    data = mean_path, aes(time, temperature),
    color = "black", linewidth = 1, linetype = "dashed", inherit.aes = FALSE
  ) +
  geom_polygon(
    data = constraint_box, aes(x, y, group = group),
    fill = NA, color = "black", linetype = "dotted", lwd = .9, inherit.aes = FALSE
  ) +
  scale_color_viridis_b(
    "Goodnes-of-fit (GOF)",
    limits = c(0, .51),
    breaks = seq(0, 1, .1),
    oob = squish
  ) +
  scale_x_reverse(position = "top") +
  scale_y_reverse() +
  coord_cartesian(expand = FALSE, ylim = c(max_temperature, 0), xlim = c(max_time, 0))
```

> The above plot uses some minor modifications of a default `ggplot2::ggplot()`, e.g. reverse the axes and colors etc. Feel free to edit these plots following your personal preferences. Check the ggplot user manual for more details: https://ggplot2.tidyverse.org/




# Clustering

First, we calculate the dissimilarities between the paths using the **Hausdorff distance**:

```{r dissimilarity}
tT_diss <- path_diss(paths_filtered, dist = "Hausdorff")
```

> An alternative measure for the dissimilarities is the Fr&#233;chet distance using `dist = 'Frechet'`.


## Is the data clustered?

Whether the data is actually clustered can be checked using the Hopkins statistic (H), i.e. the mean nearest neighbor distance in the random data set divided by the sum of the mean nearest
neighbor distances in the real and across the simulated data set

How to interpret the Hopkins statistics? If data D were uniformly
distributed, then the distances between the reals and the simulated data would be close to each other, and thus H would be about 0.5. 
However, if clusters are present in D, then
the distances for artificial points would be substantially
larger than for the real ones in expectation, and thus the
value of H will increase (Han, Kamber, and Pei 2012).

A value for H higher than 0.75 indicates a clustering tendency at the
90% confidence level.

The null and the alternative hypotheses are defined as follow:

-    Null hypothesis: the data set D is uniformly distributed (i.e., no
    meaningful clusters)

-   Alternative hypothesis: the data set D is not uniformly distributed
    (i.e., contains meaningful clusters)



> We can conduct the Hopkins Statistic test iteratively, using 0.5 as the threshold to reject the alternative hypothesis. That is, if H < 0.5, then it is unlikely that D has statistically significant clusters. 
Put in other words, If the value of Hopkins statistic is close to 1, then we can reject the null hypothesis and conclude that the dataset D is significantly a clusterable data.

```{r hopkins}
tT_diss$hopkins
```

The result shows that our data is likely clusterable.


## How many clusters?
The **optimal number of clusters** can be estimated using `path_nbclust()` and is based on the *average silhouette width*:

```{r nbcl}
path_nbclust(tT_diss)
```

The method recommends to cluster for 2 path families.

## Cluster the data

Finally we cluster the data using our estimate for the optimal number of clusters. 
The clustering is done using `cluster_paths()`, which returns a data.frame containing the path number and its assigned cluster:
```{r cluster1}
cluster_res <- paths_filtered |>
  cluster_paths(k = 2, method = "hclust")

# print the first rows of the cluster result:
head(cluster_res)
```

You can quickly check how many paths are assigned to each cluster:

```{r count_cluster}
count_cluster(cluster_res)
```

In our example, the paths are almost equally distributed in two clusters.

We can now merge the cluster result with the original path data.frame: 

```{r cluster2}
path_families <- right_join(cluster_res, paths_filtered$paths, 
                            join_by(segment))
```

We can calculate also some path statistics (e.g. median paths, 5% and 95% quantiles) for each cluster:

```{r cluster3}
path_families_binned <- path_families |> 
  group_by(cluster) |> 
  densify_cluster() |>
  path_statistics()

head(path_families_binned)
```


Plot the cluster of the modeled t-T paths:

```{r cluster_plot, fig.dim = c(7, 5)}
ggplot(data = path_families_binned) +
  geom_ribbon(
    aes(x = time_median, ymin = temp_5, ymax = temp_95, group = cluster, fill = cluster),
    stat = "stepribbon",
    inherit.aes = FALSE,
    alpha = .5
  ) +
  geom_step(
    aes(x = time_median, y = temp_median, color = cluster),
    lwd = 1.5
  ) +
  scale_color_viridis_d(aesthetics = c("fill", "color")) +

  # Axes settings:
  scale_y_reverse() +
  scale_x_reverse(position = "top") +

  # Labels:
  labs(
    title = "Example",
    subtitle = "Path families",
    x = "Time (Ma)",
    y = bquote("Temperature (" * degree * "C)")
  ) +
  # facet_wrap(vars(cluster)) + # uncomment if you want each cluster in a separate plot
  coord_cartesian(expand = FALSE, ylim = c(max_temperature, 0), xlim = c(max_time, 0))
```
