---
title: "Path density plots"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#output: pdf_document
#latex_engine: xelatex 
vignette: >
  %\VignetteIndexEntry{Path density plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(thermochron)
```

This document provides step-by-step instructions for producing density
plots derived from HeFTy inverse thermal history models as seen in
Padgett et al. (2025) and Johns-Buss et al. (2025). 

### Load Input Data

Open R and install and load the necessary packages. You can install the
packages by running the following code:

```{r packages1, eval=FALSE, include=TRUE}
install.packages("ggplot2")
remotes::install_github("tobiste/thermochron")
```


Next, install the {thermochron} package by running the following code:

```{r packages2, include=TRUE,eval=TRUE}
library(ggplot2)
library(thermochron)
```

Define the path to your Hefty output (.txt file). For example:

```{r fname}
path2myfile <- "inst/112-9-30-zr-inv.txt"
```

> Be aware that R uses forward-slashes (/) to separate folders.


Next, you import the .txt file into R by using the function
`read_hefty()`:

```{r data, eval=FALSE, echo=FALSE, include=TRUE}
tT_paths <- read_hefty(path2myfile)$paths
```

```{r data1, eval=TRUE, include=FALSE}
data("tT_paths1")
tT_paths <- tT_paths1
rm(tT_paths1)
```

## Part II

### Plot the path density

To plot the density of the paths, you simply use the function
`plot_path_density()`:

```{r seed1, include=FALSE}
set.seed(20250411)
```

```{r plot1, out.width="100%"}
plot1 <- plot_path_density_filled(tT_paths, show.legend = FALSE)
print(plot1)
```

This uses the package's default values for smoothing and binning and
creates a `ggplot` type graphic.

You can customize the smoothing and density binning by changing the
parameters - `bins` - the number of filled contours.

-   `GOF_rank` - Selects only the n highest GOF ranked paths.

-   `densify` - Should extra points be added along the individual paths
    to avoid that only the vertices of the path are evaluated?

-   `n` - How many equally-spaced extra points should be added along
    between the vertices of the path (if `densify=TRUE`).

-   `samples` - Size of a random subsample of all the paths to reduce
    the computation time.

```{r seed2, include=FALSE}
set.seed(20250411)
```

```{r plot2, out.width="100%"}
plot2 <- plot_path_density_filled(tT_paths, bins = 25, GOF_rank = 5, densify = TRUE, n = 100, max_distance = 1, samples = 100, show.legend = FALSE)
print(plot2)
```

Finally, you can customize your ggplot, such as axes labels, change
colors, and reverse the axes:

```{r plot3, out.width="100%"}
plot2 +
  labs(
    title = "Kernel density of t-T paths",
    caption = "data from Fontes Pinto et al. (in prep.)",
    x = "Time (Ma)",
    y = bquote("Temperature (" * degree * "C)")
  ) +
  theme_classic() +
  coord_cartesian(expand = FALSE) +
  scale_x_continuous(transform = "reverse", position = "top") +
  scale_y_continuous(transform = "reverse") +
  guides(fill = "none")
```
